On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2019-05-07 manuscripts/IEEE/final_submission.tex
page 1 steam rolled

\begin{abstract}
Learning optimal dictionaries for sparse coding has exposed characteristic sparse features of many natural signals. However, universal conditions guaranteeing the stability of these learned features in the presence of measurement noise have yet to be described. We prove very generally that optimal dictionaries and sparse codes are unique up to measurement error for all diverse enough datasets generated by the sparse coding model. Applications are given to data analysis, engineering, and neuroscience. 
\end{abstract}

@@

\section{Introduction}\label{Intro}
%\IEEEPARstart{L}{earning} sparse representations of data has become an important step in modern solutions to problems in signal processing and pattern analysis. 
% \cite{sato1975method}
\IEEEPARstart{A}{}common modern approach to problems in signal processing and pattern analysis is to view each of $N$ observed $n$-dimensional signal samples as a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown ``dictionary" of size $m \ll N$ (see \cite{Zhang15} for a comprehensive review of this and related models). 
Optimizing dictionaries subject to this and related sparsity constraints has revealed seemingly characteristic sparse structure in several signal classes of current interest (e.g., in vision \cite{wang2015sparse}). 
Of particular note are the seminal works in the field \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent}, which demonstrated that such dictionaries optimized with respect to small patches of ``natural" images share qualitative similarities with linear filters fit to the response properties of simple-cell neurons in mammalian visual cortex. Curiously, these waveforms (e.g., ``Gabor'' wavelets) appear in dictionaries learned by a variety of sparse coding algorithms trained over different natural image datasets; as such, it appears these latent features may be in some sense canonical \cite{donoho2001can}.

Motivated by these discoveries and more recent work proposing a theory of information transmission through random wiring bottlenecks in the brain \cite{Isely10}, linked to the field of ``compressed sensing", we address when optimal dictionaries for sparse representation are indeed identifiable from data. Answers to this question may also have implications in practice wherever an appeal is made to the latent sparse structure of data (e.g. forgery detection \cite{hughes2010, olshausen2010applied}; brain recordings \cite{jung2001imaging, agarwal2014spatially, lee2016sparse}; and gene expression \cite{wu2016stability}). While several algorithms which provably recover unique latent dictionaries under specific conditions have recently been proposed (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art), few theorems can be invoked to justify the consistency of inference under this model of data more broadly. To our knowledge, a universal guarantee of the uniqueness and stability of learned dictionaries and the sparse representations they induce over noisy data has yet to appear in the literature.
% in justifying the identification of forged paintings by their anomalous sparse representations \cite{hughes2010},

Here, we prove very generally that uniqueness and stability is a typical property of the dictionary learning problem described above. More specifically,  we show that dictionaries injective on a sparse domain are identifiable from \mbox{$N = m(k-1){m \choose k} + m$} noisy sparse linear combinations of their columns up to an error that is linear in the noise (Thm.~\ref{DeterministicUniquenessTheorem}). In fact, provided $n \geq \min(2k,m)$, in almost all cases the problem is well-posed (as per Hadamard \cite{Hadamard1902}) given enough data (Cor.~\ref{ProbabilisticCor}). These guarantees also hold for a related (and perhaps more commonly posed) optimization problem which seeks a dictionary minimizing the average number of elementary waveforms required to reconstruct each sample of the dataset. Importantly, in both cases our explicit, universal guarantees apply without the imposition of \emph{any} constraints on learned dictionaries  (e.g. that they, too, be injective over some domain) beyond an upper bound on dictionary size, which is necessary in any case to avoid a trivial solution (e.g. allowing for $m = N$ elementary waveforms). %That is, every pair of solutions of comparable size has some number of dictionary elements in common (up to noise), and similarly so for the coefficients of sparse codes they induce.

To state things precisely, let $\mathbf{A} \in \mathbb R^{n \times m}$ be a matrix with columns $\mathbf{A}_j$ ($j = 1,\ldots,m$) and let dataset $Z$ consist of measurements:
\begin{align}\label{LinearModel}
\mathbf{z}_i = \mathbf{A}\mathbf{x}_i + \mathbf{n}_i,\ \ \  \text{$i=1,\ldots,N$},
\end{align}
for $k$-\emph{sparse} $\mathbf{x}_i \in \mathbb{R}^m$ having at most $k<m$ nonzero entries and \emph{noise} $\mathbf{n}_i \in \mathbb{R}^n$, with bounded norm $\| \mathbf{n}_i \|_2 \leq  \eta$ representing our worst-case uncertainty in measuring the product $\mathbf{A}\mathbf{x}_i$. The first mathematical problem we consider is:

@@

Note that as stated, every solution to this problem represents infinitely many equivalent alternatives $\mathbf{BPD}$ and $\mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\overline x}_1, \ldots, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\overline x}_N$ parametrized by a choice of permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$. Identifying these alternatives gives a set of elementary waveforms (the columns of $\mathbf{B}$)  up to some arbitrary scale and their associated sparse coefficients (the entries of $\mathbf{x}_i$) for reconstructing each data point $\mathbf{z}_i$. Previous theoretical work addressing the noiseless case $\eta =0$ (e.g., \cite{li2004analysis, Georgiev05, Aharon06, Hillar15}) for matrices $\mathbf{B}$ having exactly $m$ columns has shown that the solution to Prob.~\ref{InverseProblem}, when it exists, is unique up to these equivalent labeling and scaling ambiguities provided the $\mathbf{x}_i$ are sufficiently diverse and the matrix $\mathbf{A}$ satisfies the \textit{spark condition}:
\begin{align}\label{SparkCondition}
\mathbf{A}\mathbf{x}_1 = \mathbf{A}\mathbf{x}_2 \implies \mathbf{x}_1 = \mathbf{x}_2, \ \ \ \text{for all $k$-sparse } \mathbf{x}_1, \mathbf{x}_2,
\end{align}
%
which in any case is necessary to guarantee the uniqueness of arbitrary $k$-sparse $\mathbf{x}_i$. We generalize these results to the practical setting  $\eta > 0$ by considering a notion of stability with respect to measurement error.

>>>>

\begin{abstract}
Learning optimal dictionaries for sparse coding has exposed characteristic sparse features of many natural signals. However, there are few universal conditions guaranteeing the stability of such features in the presence of measurement noise. Here, we prove very generally that optimal dictionaries and sparse codes are unique up to measurement error for all diverse enough datasets generated by the sparse coding model. Applications are given to data analysis, engineering, and neuroscience. 
\end{abstract}

@@

\section{Introduction}\label{Intro}
%\IEEEPARstart{L}{earning} sparse representations of data has become an important step in modern solutions to problems in signal processing and pattern analysis. 
% \cite{sato1975method}
\IEEEPARstart{A}{}common modern approach to pattern analysis in signal processing is to view each of $N$ observed $n$-dimensional signal samples as a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown ``dictionary" of size $m \ll N$ (see \cite{Zhang15} for a comprehensive review). 
Optimizing dictionaries subject to this and related sparsity constraints has revealed seemingly characteristic sparse structure in several signal classes of current interest (e.g., in vision \cite{wang2015sparse}). 
Of particular note are the seminal works in the field \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent}, which discovered that dictionaries optimized for coding small patches of ``natural" images share qualitative similarities with linear filters estimated from response properties of simple-cell neurons in mammalian visual cortex. Curiously, these waveforms (e.g., ``Gabor'' wavelets) appear in dictionaries learned by a variety of algorithms trained over different natural image datasets; in particular, sparse coding of  latent features in natural signals may, in some sense, be canonical \cite{donoho2001can}.

Motivated by these discoveries and more recent work relating ``compressed sensing" to a theory of information transmission through random wiring bottlenecks in the brain \cite{Isely10}, we address when optimal dictionaries for sparse representation are indeed identifiable from data. Answers to this question may also have implications in practice wherever an appeal is made to latent sparse structure of data (e.g., forgery detection \cite{hughes2010, olshausen2010applied}; brain recordings \cite{jung2001imaging, agarwal2014spatially, lee2016sparse}; and gene expression \cite{wu2016stability}). While several algorithms have recently been proposed that provably recover unique dictionaries under specific conditions (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art), few theorems can be invoked to justify the consistency of inference under this model of data more broadly. To our knowledge, a universal guarantee of the uniqueness and stability of learned dictionaries and the sparse representations they induce when data is noisy has yet to appear in the literature.
% in justifying the identification of forged paintings by their anomalous sparse representations \cite{hughes2010},

Here, we prove very generally that uniqueness and stability is a typical property of sparse dictionary learning. More specifically,  we show that dictionaries injective on a sparse domain are identifiable from \mbox{$N = m(k-1){m \choose k} + m$} noisy sparse linear combinations of their columns up to an error that is linear in the noise (Thm.~\ref{DeterministicUniquenessTheorem}). In fact, provided $n \geq \min(2k,m)$, in almost all cases the problem is well-posed (as per Hadamard \cite{Hadamard1902}) given enough data (Cor.~\ref{ProbabilisticCor}). These guarantees also hold for a related (and perhaps more commonly posed) optimization problem which seeks a dictionary minimizing the average number of elementary waveforms required to reconstruct each sample of the dataset (Thm.~\ref{SLCopt}). Importantly, in both cases our explicit, universal guarantees apply without the imposition of \emph{any} constraints on learned dictionaries  (e.g. that they, too, be injective over some domain) beyond an upper bound on dictionary size, which is necessary in any case to avoid a trivial solution (e.g., allowing for $m = N$ elementary waveforms). %That is, every pair of solutions of comparable size has some number of dictionary elements in common (up to noise), and similarly so for the coefficients of sparse codes they induce.

More precisely, let $\mathbf{A} \in \mathbb R^{n \times m}$ be a matrix with columns $\mathbf{A}_j$ ($j = 1,\ldots,m$) and let dataset $Z$ consist of measurements:
\begin{align}\label{LinearModel}
\mathbf{z}_i = \mathbf{A}\mathbf{x}_i + \mathbf{n}_i,\ \ \  \text{$i=1,\ldots,N$},
\end{align}
for $k$-\emph{sparse} $\mathbf{x}_i \in \mathbb{R}^m$ having at most $k<m$ nonzero entries and \emph{noise} $\mathbf{n}_i \in \mathbb{R}^n$, with bounded norm $\| \mathbf{n}_i \|_2 \leq  \eta$ representing our worst-case uncertainty in measuring the product $\mathbf{A}\mathbf{x}_i$. The first mathematical problem we consider is the following.

@@

Note that every solution to Prob.~\ref{InverseProblem} represents infinitely many equivalent alternatives $\mathbf{BPD}$ and $\mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\overline x}_1, \ldots, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\overline x}_N$ parametrized by a choice of permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$. Identifying these ambiguities (labelling and scale) yields a set of elementary waveforms (the columns of $\mathbf{B}$) and their associated sparse coefficients (the entries of $\mathbf{\overline x}_i$) for reconstructing each data point $\mathbf{z}_i$.

Previous theoretical work addressing the noiseless case $\eta =0$ (e.g., \cite{li2004analysis, Georgiev05, Aharon06, Hillar15}) for matrices $\mathbf{B}$ having exactly $m$ columns has shown that a solution to Prob.~\ref{InverseProblem}, when it exists, is unique up to such relabeling and rescaling provided the $\mathbf{x}_i$ are sufficiently diverse and $\mathbf{A}$ satisfies the \textit{spark condition}:
\begin{align}\label{SparkCondition}
\mathbf{A}\mathbf{x}_1 = \mathbf{A}\mathbf{x}_2 \implies \mathbf{x}_1 = \mathbf{x}_2, \ \ \ \text{for all $k$-sparse } \mathbf{x}_1, \mathbf{x}_2,
\end{align}
%
which in any case is necessary to guarantee the uniqueness of arbitrary $k$-sparse $\mathbf{x}_i$. We generalize these results to the practical setting  $\eta > 0$ by considering a natural notion of stability with respect to measurement error.
